---
title: "Detecting Circadian Rhythmicity in DNA Methylation with cirBEM"
author:
  - name: "AUTHOR_FIRST AUTHOR_LAST"
    affiliation: "Institution"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{cirBEM Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    message = FALSE,
    warning = FALSE
)
```

# Introduction

cirBEM is an R package for detecting
circadian rhythmicity in DNA methylation data. The package fits beta regression
models with cosinor terms to account for the bounded nature of methylation beta
values and supports repeated-measures designs through subject-specific random
effects. To improve stability in genome-wide analyses, cirBEM also incorporates
empirical Bayesian shrinkage of the beta-regression precision parameter.

This vignette demonstrates how to use cirBEM for:
1. Fitting beta-regression cosinor models to methylation data.
2. Detecting rhythmic CpG sites and ranking results.
3. Visualizing fitted circadian methylation patterns.

# Installation

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("cirBEM")
```

# Quick Start

```{r quickstart, eval=FALSE}
library(cirBEM)

# Start from a SummarizedExperiment with methylation beta values
se <- your_summarized_experiment

# Fit with repeated measures and EB shrinkage
se <- fitCosinor(se, time_col = "time", subject_col = "id")

# View top rhythmic CpGs
topCpGs(se, n = 10)

# Visualize
plotCosinorFit(se)
```

# Data Preparation

cirBEM expects a `SummarizedExperiment` with:

- An assay containing methylation beta values (CpGs x samples)
- `colData` with at minimum a numeric time column (in hours)
- Optionally, a subject identifier column for repeated measures

The package includes example data (3000 CpGs, 8 subjects, 6 timepoints):

```{r data-prep}
library(cirBEM)
library(SummarizedExperiment)

# Load bundled example data
example <- readRDS(system.file("extdata", "example_data.rds",
    package = "cirBEM"))
se <- example$se
se
```

For a fast vignette build, the fitted examples below use a small subset of the
bundled dataset.

```{r data-subset}
se_demo <- se[seq_len(100), ]
```

# Cross-Sectional Analysis (Fixed-Effects)

When each subject is measured at a single timepoint:

```{r fixed, eval=FALSE}
# Use one timepoint per subject for demonstration, with collection times
# staggered across subjects so the fixed-effects model has time variation.
cd <- as.data.frame(colData(se_demo))
subjects <- unique(cd$subject)
cross_idx <- unlist(lapply(seq_along(subjects), function(i) {
    idx <- which(cd$subject == subjects[i])
    idx[((i - 1) %% length(idx)) + 1]
}))
se_cross <- se_demo[, cross_idx]

se_cross <- fitCosinor(se_cross,
    time_col = "time",
    subject_col = NULL,  # no repeated measures
    eb_shrink = TRUE,
    min_cpgs = 50,
    BPPARAM = BiocParallel::SerialParam(),
    verbose = TRUE
)
se_cross
```

# Longitudinal Analysis (Mixed-Effects)

When subjects have repeated measurements across time:

```{r mixed}
se_demo <- fitCosinor(se_demo,
    time_col = "time",
    subject_col = "subject",
    eb_shrink = TRUE,
    min_cpgs = 50,
    BPPARAM = BiocParallel::SerialParam(),
    verbose = TRUE
)

```

# Interpreting Results

All results are stored in the `SummarizedExperiment`:

- `rowData(se_demo)` - per-CpG results (coefficients, circadian
  parameters, test statistics, convergence)
- `metadata(se_demo)$cirBEM` - design info, EB prior, test settings, and fit summary

## Top CpGs

```{r top-cpgs}
topCpGs(se_demo, n = 10)
```

## Circadian Parameters

All per-CpG results are appended to `rowData(se_demo)`:

In mixed-effects mode, response-scale amplitude and MESOR are
population-average summaries: the fitted curve is averaged
over the random subject intercept distribution rather than evaluated at
`u = 0`.

```{r params}
res <- rowData(se_demo)

# Amplitude on the probability scale
head(res$amplitude_response)

# Phase in hours
head(res$phase)

# MESOR (rhythm-adjusted mean)
head(res$mesor_response)
```

# Visualization

```{r plots, fig.width=8, fig.height=5}
# Fitted curves for top CpGs
plotCosinorFit(se_demo)
```

# Session Info

```{r session}
sessionInfo()
```
