---
title: "Detecting Circadian Rhythmicity in DNA Methylation with cirBEM"
author:
  - name: "AUTHOR_FIRST AUTHOR_LAST"
    affiliation: "Institution"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{cirBEM Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    message = FALSE,
    warning = FALSE
)
```

# Introduction

DNA methylation beta values are bounded in (0, 1), making standard
linear models inappropriate for modeling circadian rhythmicity at CpG
sites. **cirBEM** addresses this by fitting beta regression models
with cosinor terms, properly accounting for the bounded nature of
methylation array data.

Key features:

- **Beta regression** with logit link for methylation proportions
- **Cosinor model** (sin/cos terms) for circadian rhythmicity
- **Fixed-effects** mode for cross-sectional designs (one sample per
  subject)
- **Mixed-effects** mode for repeated measures designs (multiple
  timepoints per subject)
- **Empirical Bayes shrinkage** of the precision parameter (phi)
  across CpG sites, analogous to limma's moderated variance

# Installation

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("cirBEM")
```

# Quick Start

```{r quickstart, eval=FALSE}
library(cirBEM)

# Load your data as a SummarizedExperiment
se <- makeCircMethylSE(beta_matrix, sample_metadata)

# Fit with repeated measures and EB shrinkage
se <- fitCosinor(se, time_col = "time", subject_col = "id")

# View top rhythmic CpGs
topCpGs(se, n = 10)

# Visualize
plotCosinorFit(se)
plotVolcano(se)
```

# Data Preparation

cirBEM expects a `SummarizedExperiment` with:

- An assay containing methylation beta values (CpGs x samples)
- `colData` with at minimum a numeric time column (in hours)
- Optionally, a subject identifier column for repeated measures

The package includes example data (100 CpGs, 8 subjects, 6 timepoints):

```{r data-prep}
library(cirBEM)
library(SummarizedExperiment)

# Load bundled example SummarizedExperiment
se <- readRDS(system.file("extdata", "example_se.rds", package = "cirBEM"))
se
```

# Cross-Sectional Analysis (Fixed-Effects)

When each subject is measured at a single timepoint:

```{r fixed, eval=FALSE}
# Use only first timepoint per subject for demonstration
first_tp <- !duplicated(colData(se)$subject)
se_cross <- se[, first_tp]

se_cross <- fitCosinor(se_cross,
    time_col = "time",
    subject_col = NULL,  # no repeated measures
    eb_shrink = TRUE,
    BPPARAM = BiocParallel::SerialParam(),
    verbose = TRUE
)
se_cross
```

# Longitudinal Analysis (Mixed-Effects)

When subjects have repeated measurements across time:

```{r mixed}
se <- fitCosinor(se,
    time_col = "time",
    subject_col = "subject",
    eb_shrink = TRUE,
    BPPARAM = BiocParallel::SerialParam(),
    verbose = TRUE
)
```

# Interpreting Results

All results are stored in the `SummarizedExperiment`:

- `rowData(se)$cirBEM` — per-CpG results (coefficients, circadian
  parameters, test statistics, convergence)
- `metadata(se)$cirBEM` — design info and EB prior

## Top CpGs

```{r top-cpgs}
topCpGs(se, n = 10)
```

## Circadian Parameters

All per-CpG results are in `rowData(se)$cirBEM`:

```{r params}
res <- rowData(se)$cirBEM

# Amplitude on the probability scale
head(res$amplitude_response)

# Acrophase in hours
head(res$acrophase_hours)

# MESOR (rhythm-adjusted mean)
head(res$mesor_response)
```

# Empirical Bayes Shrinkage

The precision parameter phi controls the variance of the beta
distribution. With small sample sizes, per-CpG estimates of phi are
noisy. Empirical Bayes shrinkage borrows strength across CpGs by
estimating a Normal prior on log(phi) and incorporating it into the
model via penalized likelihood.

```{r eb-shrinkage}
# View the estimated prior
metadata(se)$cirBEM$ebPrior

# Visualize shrinkage
plotPrecisionShrinkage(se)
```

# Visualization

```{r plots, fig.width=8, fig.height=5}
# Fitted curves for top CpGs
plotCosinorFit(se)

# Volcano plot
plotVolcano(se)
```

# Using Wald Test

By default, a likelihood ratio test is used. You can also apply a
Wald test:

```{r wald}
se_wald <- testRhythmicity(se, method = "Wald")
topCpGs(se_wald, n = 5)
```

# Session Info

```{r session}
sessionInfo()
```
