---
title: "Detecting Circadian Rhythmicity in DNA Methylation with CirBEM"
date: ""
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{CirBEM Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = FALSE,
    comment = "#>",
    message = FALSE,
    warning = FALSE
)
```

# Introduction

CirBEM is an R package for detecting circadian rhythmicity in DNA methylation
data. DNA methylation beta values are bounded between 0 and 1 and often exhibit
strong mean-variance dependence, making standard Gaussian cosinor models less
well suited for modeling methylation dynamics. In addition, circadian
methylation studies commonly involve repeated measurements from the same
subjects, requiring methods that can account for within-subject correlation. Existing rhythmicity detection methods are often designed for approximately Gaussian gene-expression measurements or independent samples. 

CirBEM addresses these issues by fitting cosinor-based beta regression models
for methylation beta values. The model represents rhythmic methylation patterns
using sine and cosine terms, while optional subject-specific random intercepts
account for repeated measurements. To improve stability in genome-wide analyses,
CirBEM also incorporates empirical Bayesian shrinkage of the beta-regression
precision parameter.

This vignette demonstrates how to use CirBEM for:

1. Fitting fixed- and mixed-effect beta-regression cosinor models to methylation
   data.
2. Detecting rhythmic CpG sites and ranking results.
3. Visualizing fitted circadian methylation patterns.

# Installation

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("cirBEM")
```

# Data Preparation

CirBEM expects a `SummarizedExperiment` with: 1. An assay containing methylation beta values (CpGs x samples); 2.`colData` with at minimum a numeric time column (in hours) 3.Optionally, a subject identifier column for repeated measures.

The package includes example data (3000 CpGs, 8 subjects, 6 timepoints):

```{r data-prep}
library(cirBEM)
library(SummarizedExperiment)

# Load bundled example data
example <- readRDS(system.file("extdata", "example_data.rds",
    package = "cirBEM"))
se <- example$se
se
```

For a fast vignette build, the fitted examples below use a small subset of the
bundled dataset.

```{r data-subset}
se_demo <- se[seq_len(100), ]
```

# Fixed-Effects Analysis

A fixed-effects model uses all samples and does not include subject-specific
random intercepts. For repeated-measures data, this treats observations as
independent and can be useful as a simple baseline model.

```{r fixed, eval=FALSE}
se_fixed <- fitCosinor(se_demo,
    time_col = "time",
    mode = "fixed",
    eb_shrink = TRUE,
    min_cpgs = 50,
    BPPARAM = BiocParallel::SerialParam(),
    verbose = TRUE
)
se_fixed
```

# Longitudinal Analysis (Mixed-Effects)

When subjects have repeated measurements across time:

```{r mixed}
se_demo <- fitCosinor(se_demo,
    time_col = "time",
    subject_col = "subject",
    eb_shrink = TRUE,
    min_cpgs = 50,
    BPPARAM = BiocParallel::SerialParam(),
    verbose = TRUE
)

```

# Interpreting Results

The output of `fitCosinor()` is the input `SummarizedExperiment` object with
model results appended to `rowData()` and analysis-level information stored in
`metadata()`.

Specifically:

- `rowData(se_demo)` contains per-CpG model estimates, circadian summary
  statistics, hypothesis test results, adjusted p-values, and convergence
  indicators.
- `metadata(se_demo)$cirBEM` contains analysis-level information, including
  the model design, empirical Bayes prior estimates, test settings, and fitting
  summary.

## Circadian parameter estimates

All per-CpG model results are available from `rowData(se_demo)`:

```{r results}
res <- rowData(se_demo)
head(res)
```

The main circadian summaries are:

- `amplitude_response`: estimated oscillation amplitude on the methylation
  beta-value scale.
- `phase`: estimated timing of peak methylation, reported in hours.
- `mesor_response`: rhythm-adjusted mean methylation level on the beta-value
  scale.

In mixed-effects mode, response-scale summaries are reported as
population-average quantities. That is, the fitted methylation curve is averaged
over the subject-specific random-intercept distribution, rather than evaluated
at a random effect of zero. This distinction is important because the inverse
logit transformation is nonlinear.

## Top-ranked CpG sites

The function `topCpGs()` returns the top-ranked CpG sites according to the
rhythmicity test p-value or adjusted p-value, depending on the available result
columns.

```{r top-cpgs}
topCpGs(se_demo, n = 10)
```

# Visualization

The fitted circadian methylation pattern can be visualized using
`plotCosinorFit()`. By default, the function displays fitted 24-hour cosinor
curves for the top-ranked CpG sites.

```{r plots, fig.width=8, fig.height=5}
plotCosinorFit(se_demo)
```

# Session Info

```{r session}
sessionInfo()
```
